---
title: "DATA 612 : Project 1"
author: "Jerald Melukkaran"
date: "2026-06-06"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

---

## 1. Introduction 

We are going to build a recommender system that predicts movie recommendantions using data gathered from a 10 by 10 matrix of users and movies. 
As per the video lectures, we start with the Raw global averages, calculate the RMSE for the test and training data set, and then make the baseline estimate adjusted with the user and movie biases and see if there are any reasonable improvements 

---

## 2. Dataset

```{r build-data}
library(tidyverse)
library(reshape2)

set.seed(3)

movies <- c("Inception", "The Matrix", "Interstellar",
            "The Dark Knight", "Parasite",
            "Avengers", "Joker", "1917",
            "Ford v Ferrari", "Knives Out")
users <- paste0("User", 1:10)

ratings_raw <- matrix(
  c( 
    5,   4,   5,   NA,  3,   4,   NA,  3,   4,   5,  
    3,   2,   NA,  3,   2,   NA,  2,   3,   NA,  2,  
    4,   4,   3,   4,   NA,  3,   4,   NA,  4,   3,   
    NA,  5,   4,   5,   4,   5,   NA,  4,   5,   NA, 
    2,   NA,  2,   3,   2,   NA,  1,   2,   3,   2,   
    4,   3,   4,   NA,  4,   3,   4,   3,   NA,  4,   
    NA,  4,   NA,  5,   3,   4,   5,   NA,  4,   5,  
    3,   3,   4,   3,   NA,  3,   NA,  4,   3,   3,  
    5,   NA,  5,   4,   5,   4,   4,   5,   NA,  4,  
    2,   3,   NA,  2,   2,   3,   2,   NA,  2,   3   
  ),
  nrow = 10, ncol = 10, byrow = TRUE,
  dimnames = list(users, movies)
)

cat("Raw User–Item Matrix:\n")
print(ratings_raw)

```

---

## 3. Splitting into Testing and Training sets 8:20 

```{r train-test-split}
ratings_long <- ratings_raw %>%
  as.data.frame() %>%
  rownames_to_column("user") %>%
  pivot_longer(-user, names_to = "item", values_to = "rating") %>%
  filter(!is.na(rating))

cat("Total observed ratings:", nrow(ratings_long), "\n")

set.seed(123)
test_idx   <- sample(seq_len(nrow(ratings_long)), size = floor(0.20 * nrow(ratings_long)))
train_data <- ratings_long[-test_idx, ]
test_data  <- ratings_long[ test_idx, ]

cat("Training ratings:", nrow(train_data), "\n")
cat("Test ratings    :", nrow(test_data),  "\n")

# Training set only: This is what we'll use for calculations 
train_matrix <- train_data %>%
  pivot_wider(names_from = item, values_from = rating) %>%
  column_to_rownames("user")

cat("\nTraining Matrix")
print(train_matrix)
```

---

## 4. Raw Average Prediction

As per the video, this is our simplest estimate, Basically an average of all available ratings in the training set 

```{r raw-average}
global_mean <- mean(train_data$rating)
cat("Global mean rating (training):", round(global_mean, 4), "\n")

train_data <- train_data %>% mutate(pred_raw = global_mean)
test_data  <- test_data  %>% mutate(pred_raw = global_mean)
```


Now we can try two advanced error estimation Methods 
age

### Root mean squared error 

Error = Observed - Raw average 

-> then Squared

-> then taken the Mean of 

-> then taken the Square root of '

### Mean absolute error 

Error = Observed - Predicted 

-> Absolute value taken 

-> Mean taken



RMSE because errors are squared, punishes large individual mistakes more heavily than small ones.MAE treats every mistake equally regardless of size so it helps us show the usual prediction error a user would actually experience.Comparing the two should give us a better idea on the spread. Ff RMSE is much larger than MAE, it means a few predictions are very far off. If they are close together, errors are spread fairly evenly across predictions.




```{r rmse-raw}
rmse <- function(actual, predicted) sqrt(mean((actual - predicted)^2))
mae  <- function(actual, predicted) mean(abs(actual - predicted))

rmse_raw_train <- rmse(train_data$rating, train_data$pred_raw)
rmse_raw_test  <- rmse(test_data$rating,  test_data$pred_raw)
mae_raw_train  <- mae(train_data$rating,  train_data$pred_raw)
mae_raw_test   <- mae(test_data$rating,   test_data$pred_raw)

cat("RMSE Raw Average – Train:", round(rmse_raw_train, 4), "\n")
cat("RMSE Raw Average – Test :", round(rmse_raw_test,  4), "\n")
cat("MAE  Raw Average – Train:", round(mae_raw_train,  4), "\n")
cat("MAE  Raw Average – Test :", round(mae_raw_test,   4), "\n")
```




## 5. Bias Calculation

Since the raw average doesent really tell us anything about the users or the movies, we need to introduce bias. This helps us find out how users rate movies, if a specific user is a harsh or easy critic or if a specific movie is generally good or bad.

Bias for User 1 = Average of all their ratings given - Raw average (Positive if User is more lenient, Negative if they are harsher)

Bias for Movie 1 = Average of all their ratings gotten - Raw average (Positive if Movie is generally good, Negative if its bad )

This helps us create the final baseline predictor 

Fina predictor = Raw average + User bias + Movie bias 

```{r biases}
# User bias
user_bias <- train_data %>%
  group_by(user) %>%
  summarise(user_mean = mean(rating), .groups = "drop") %>%
  mutate(b_u = user_mean - global_mean)

cat("User Biases:\n")
print(user_bias %>% select(user, user_mean, b_u) %>% arrange(user))

# Item bias
item_bias <- train_data %>%
  left_join(user_bias %>% select(user, b_u), by = "user") %>%
  mutate(residual = rating - global_mean - b_u) %>%
  group_by(item) %>%
  summarise(b_i = mean(residual), .groups = "drop")

cat("\nItem Biases:\n")
print(item_bias %>% arrange(desc(b_i)))
```



---

## 6. Baseline Predictor


```{r baseline-predictor}
# Join biases into training set
train_data <- train_data %>%
  left_join(user_bias %>% select(user, b_u), by = "user") %>%
  left_join(item_bias,                        by = "item") %>%
  mutate(pred_baseline = pmax(1, pmin(5, global_mean + b_u + b_i)))

# Join biases into test set
test_data <- test_data %>%
  left_join(user_bias %>% select(user, b_u), by = "user") %>%
  left_join(item_bias,                        by = "item") %>%
  mutate(pred_baseline = pmax(1, pmin(5, global_mean + b_u + b_i)))

cat("Sample baseline predictions (test set):\n")
test_data %>%
  select(user, item, rating, pred_raw, pred_baseline) %>%
  slice_head(n = 10) %>%
  print()
```

---

## 7. Full Predicted Matrix

```{r full-matrix}
# All user–item combinations
full_grid <- expand.grid(user = users, item = movies, stringsAsFactors = FALSE)

full_preds <- full_grid %>%
  left_join(user_bias %>% select(user, b_u), by = "user") %>%
  left_join(item_bias,                        by = "item") %>%
  mutate(pred_baseline = pmax(1, pmin(5, global_mean + b_u + b_i))) %>%
  select(user, item, pred_baseline)

pred_matrix <- full_preds %>%
  pivot_wider(names_from = item, values_from = pred_baseline) %>%
  column_to_rownames("user")

cat("Full Baseline Prediction Matrix (all user–item pairs):\n")
print(round(pred_matrix, 2))
```

---

## 8. RMSE & MAE – Baseline Predictor

```{r rmse-baseline}
rmse_bl_train <- rmse(train_data$rating, train_data$pred_baseline)
rmse_bl_test  <- rmse(test_data$rating,  test_data$pred_baseline)
mae_bl_train  <- mae(train_data$rating,  train_data$pred_baseline)
mae_bl_test   <- mae(test_data$rating,   test_data$pred_baseline)

cat("RMSE Baseline – Train:", round(rmse_bl_train, 4), "\n")
cat("RMSE Baseline – Test :", round(rmse_bl_test,  4), "\n")
cat("MAE  Baseline – Train:", round(mae_bl_train,  4), "\n")
cat("MAE  Baseline – Test :", round(mae_bl_test,   4), "\n")
```

---

## 9. Results

```{r summary-table}
results <- tibble(
  Method  = c("Raw Average", "Raw Average", "Baseline Predictor", "Baseline Predictor"),
  Dataset = c("Train", "Test", "Train", "Test"),
  RMSE    = c(rmse_raw_train, rmse_raw_test, rmse_bl_train, rmse_bl_test),
  MAE     = c(mae_raw_train,  mae_raw_test,  mae_bl_train,  mae_bl_test)
)

results %>%
  mutate(RMSE = round(RMSE, 4), MAE = round(MAE, 4)) %>%
  knitr::kable(caption = "RMSE & MAE: Raw Average vs Baseline Predictor")
```

---

## 10. Interpretation & Conclusions


Bias adjustment meaningfully lowers both RMSE and MAE when compared to using Raw averages. 

When we start to account for the fact that some Users are more generous (User1, User4, User9) and some Movies are better loved (The Dark Knight, Inception), the baseline predictor reduces prediction error on the test set. 

The gap between training and test set in both metrics is also relatively small, so our method estimates generalise well.

**RMSE vs MAE**

RMSE is always going to be greater than or equal to MAE because squaring amplifies larger errors. The ratio between them tells us about the shape of our errors. Here RMSE is consistently about 0.10–0.15 higher than MAE across every case. That small, stable gap tells us the errors in this model are fairly uniform in size there are no wild outlier predictions dragging RMSE up disproportionaly. Also This suggests our extreme users the very generous and very harsh raters are still producing some bigger errors that bias correction did not fully eliminate.

The ratio between RMSE and AME stays stable even after adding bias correction, meaning the baseline did not introduce any new extreme wrong predictions, it just shifted all errors down uniformly.

Baseline predictor brings MAE down from 0.90 to 0.51 on the test set, meaning on average we are now off by about half a point instead of nearly a full point. That is a meaningful improvement

MAE is easier to communicate , while RMSE is more useful as maybe a safeguard against like recommending a 1 star movie to someone who would probably rate it a 5. Since our RMSE and MAE stay proportional throughout, we can be reasonably confident the model is not producing those kinds of outlier recommendations.
