---
title: "DATA 612 – Project 2: Collaborative Filtering on the Jester Dataset"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This project implements two collaborative filtering recommender systems on the Jester Joke Dataset — a collection of joke ratings (−10 to +10) from real users. The two approaches are **User-User Collaborative Filtering**, which finds users who rated jokes similarly to the active user and borrows their ratings to make predictions, and **Item-Item Collaborative Filtering**, which finds jokes that tend to get rated similarly to each other and uses that to predict missing ratings.

Both algorithms are evaluated using RMSE, MAE, and NMAE on a held-out test set. We also try both Pearson and Cosine similarity to see which performs better.

---

## Load Libraries

```{r libraries}
library(readxl)    # reading the .xls file
library(dplyr)     # data wrangling
library(ggplot2)   # plotting
library(tidyr)     # reshaping for plots
```

---

## Load and Clean the Data

```{r load-data}
# Read the Jester data file. It has 101 columns:
#   Column 1: how many jokes that user rated (we don't need this)
#   Columns 2-101: the user's rating for jokes 1-100
# Ratings go from -10 to +10. The value 99 means "not rated" — we replace it with NA.

raw <- read_excel("jester-data-1.xls", col_names = FALSE)

# Drop the first column (rating count) and name the rest
ratings <- raw[, -1]
colnames(ratings) <- paste0("joke", 1:100)

# Replace 99 (the "not rated" sentinel) with NA
ratings[ratings == 99] <- NA

# Convert to a plain matrix for easier math
ratings_mat <- as.matrix(ratings)

cat("Dataset size:", nrow(ratings_mat), "users x", ncol(ratings_mat), "jokes\n")
cat("Percentage of ratings that are observed:",
    round(100 * mean(!is.na(ratings_mat)), 1), "%\n")
```

---

## Sample and Split

```{r sample-split}
set.seed(612)

# The full dataset has ~25,000 users. We sample 1,000 to keep runtime reasonable.
# We only keep users who rated at least 20 jokes so similarity estimates are meaningful.
enough_ratings <- which(rowSums(!is.na(ratings_mat)) >= 20)
sampled        <- sample(enough_ratings, 1000)
mat            <- ratings_mat[sampled, ]

# Train/Test split: hold out 20% of the observed ratings for evaluation.
# We'll predict those held-out values and compare to ground truth.
observed    <- which(!is.na(mat), arr.ind = TRUE)
test_idx    <- sample(nrow(observed), size = round(0.2 * nrow(observed)))

train_mat   <- mat
test_mat    <- matrix(NA, nrow = nrow(mat), ncol = ncol(mat))

# Move the test ratings out of the training matrix
for (i in test_idx) {
  r <- observed[i, 1]
  c <- observed[i, 2]
  test_mat[r, c]  <- mat[r, c]
  train_mat[r, c] <- NA
}

cat("Training ratings:", sum(!is.na(train_mat)), "\n")
cat("Test ratings:    ", sum(!is.na(test_mat)),  "\n")
```

---

## Error Metric Functions

```{r metrics}
# We use three metrics to evaluate predictions.
#
# RMSE (Root Mean Squared Error): the most common metric. It penalizes
#   big errors more heavily than small ones. Lower is better.
#
# MAE (Mean Absolute Error): the average size of the error in rating points.
#   Easier to interpret — an MAE of 3 means predictions are off by 3 points
#   on average on a scale of -10 to +10.
#
# NMAE (Normalized MAE): MAE divided by the rating range (20 points).
#   This turns MAE into a percentage. The Jester paper reports ~20% NMAE
#   for a good recommender; random guessing is about 33%.

rmse <- function(actual, pred) sqrt(mean((actual - pred)^2, na.rm = TRUE))
mae  <- function(actual, pred) mean(abs(actual - pred), na.rm = TRUE)
nmae <- function(actual, pred) mae(actual, pred) / 20  # range is 20 (-10 to +10)
```

---

## Similarity Functions

```{r similarity}
# Pearson similarity: measures how correlated two rating vectors are.
# It naturally handles the fact that some users rate everything high
# and others rate everything low, because correlation is about the
# *pattern*, not the absolute level.
pearson_sim <- function(a, b) {
  idx <- !is.na(a) & !is.na(b)
  if (sum(idx) < 2) return(0)   # need at least 2 shared ratings
  cor(a[idx], b[idx])
}

# Cosine similarity: treats each rating vector as a point in space
# and measures the angle between them. Does not correct for rating
# level bias the way Pearson does.
cosine_sim <- function(a, b) {
  idx <- !is.na(a) & !is.na(b)
  if (sum(idx) == 0) return(0)
  a2 <- a[idx]; b2 <- b[idx]
  denom <- sqrt(sum(a2^2)) * sqrt(sum(b2^2))
  if (denom == 0) return(0)
  sum(a2 * b2) / denom
}
```

---

## User-User Collaborative Filtering

```{r uucf}
# USER-USER COLLABORATIVE FILTERING
#
# The idea: to predict what user A would think of a joke they haven't
# rated, find other users who have similar tastes (similar rating patterns),
# then take a weighted average of what those similar users thought of that joke.
#
# Steps:
#   1. Compute similarity between the target user and all other users
#   2. Pick the top k most similar users (neighbors) who have rated the target joke
#   3. Predict = weighted average of neighbor ratings (weights = similarities)
#
# We mean-center ratings before computing similarity. This removes the
# effect of users who tend to rate everything high or low.

# Pre-compute each user's average rating
user_means <- rowMeans(train_mat, na.rm = TRUE)

# Mean-center the training matrix
train_centered <- sweep(train_mat, 1, user_means, FUN = "-")
train_centered[is.na(train_mat)] <- NA

uucf_predict <- function(sim_fn, k = 15) {
  # Find all test cells (user, joke pairs we need to predict)
  test_cells <- which(!is.na(test_mat), arr.ind = TRUE)
  
  actual    <- numeric(nrow(test_cells))
  predicted <- numeric(nrow(test_cells))
  
  for (i in seq_len(nrow(test_cells))) {
    u <- test_cells[i, 1]   # target user
    j <- test_cells[i, 2]   # target joke
    
    actual[i] <- test_mat[u, j]
    
    # Compute similarity between target user and all other users
    sims <- sapply(seq_len(nrow(train_mat)), function(v) {
      if (v == u) return(NA)
      sim_fn(train_centered[u, ], train_centered[v, ])
    })
    
    # Only keep neighbors who actually rated joke j
    has_rated <- !is.na(train_mat[, j])
    sims[!has_rated] <- NA
    sims[u] <- NA
    
    # Pick top k neighbors by similarity
    top_k <- order(sims, decreasing = TRUE, na.last = NA)[1:min(k, sum(!is.na(sims)))]
    
    if (length(top_k) == 0) {
      predicted[i] <- user_means[u]   # fall back to user's own mean
      next
    }
    
    w <- sims[top_k]
    if (sum(abs(w)) == 0) {
      predicted[i] <- user_means[u]
      next
    }
    
    # Weighted average of centered neighbor ratings + target user's mean
    neighbor_ratings <- train_centered[top_k, j]
    pred <- user_means[u] + sum(w * neighbor_ratings, na.rm = TRUE) / sum(abs(w))
    predicted[i] <- max(-10, min(10, pred))   # clip to valid range
  }
  
  data.frame(actual = actual, predicted = predicted)
}

cat("Running UUCF with Pearson similarity (k=15)...\n")
uucf_pearson <- uucf_predict(pearson_sim, k = 15)

cat("Running UUCF with Cosine similarity (k=15)...\n")
uucf_cosine  <- uucf_predict(cosine_sim,  k = 15)
```

---

## Item-Item Collaborative Filtering

```{r iicf}
# ITEM-ITEM COLLABORATIVE FILTERING
#
# Instead of comparing users, we compare jokes. The logic: if two jokes
# tend to get similar ratings from the same people, they are "similar."
# To predict what a user thinks of a joke they haven't rated, we find
# the most similar jokes that the user *has* rated and average those.
#
# Item-item CF is often more stable than user-user because item
# relationships change less over time than user preferences.

iicf_predict <- function(sim_fn, k = 15) {
  test_cells <- which(!is.na(test_mat), arr.ind = TRUE)
  
  actual    <- numeric(nrow(test_cells))
  predicted <- numeric(nrow(test_cells))
  
  for (i in seq_len(nrow(test_cells))) {
    u <- test_cells[i, 1]
    j <- test_cells[i, 2]
    
    actual[i] <- test_mat[u, j]
    
    # Find jokes the user has rated (in training data)
    rated_by_user <- which(!is.na(train_mat[u, ]))
    rated_by_user <- setdiff(rated_by_user, j)
    
    if (length(rated_by_user) == 0) {
      predicted[i] <- user_means[u]
      next
    }
    
    # Similarity between target joke j and all jokes user has rated
    sims <- sapply(rated_by_user, function(jj) sim_fn(train_mat[, j], train_mat[, jj]))
    
    # Top k most similar jokes
    top_k <- order(sims, decreasing = TRUE, na.last = NA)[1:min(k, length(sims))]
    w     <- sims[top_k]
    
    if (sum(abs(w)) == 0) {
      predicted[i] <- user_means[u]
      next
    }
    
    neighbor_jokes   <- rated_by_user[top_k]
    neighbor_ratings <- train_mat[u, neighbor_jokes]
    pred <- sum(w * neighbor_ratings, na.rm = TRUE) / sum(abs(w))
    predicted[i] <- max(-10, min(10, pred))
  }
  
  data.frame(actual = actual, predicted = predicted)
}

cat("Running IICF with Pearson similarity (k=15)...\n")
iicf_pearson <- iicf_predict(pearson_sim, k = 15)

cat("Running IICF with Cosine similarity (k=15)...\n")
iicf_cosine  <- iicf_predict(cosine_sim,  k = 15)
```

---

## Results

```{r results-table}
# Collect all results into one comparison table
results <- data.frame(
  Algorithm  = c("UUCF – Pearson", "UUCF – Cosine",
                 "IICF – Pearson", "IICF – Cosine"),
  RMSE = c(rmse(uucf_pearson$actual, uucf_pearson$predicted),
           rmse(uucf_cosine$actual,  uucf_cosine$predicted),
           rmse(iicf_pearson$actual, iicf_pearson$predicted),
           rmse(iicf_cosine$actual,  iicf_cosine$predicted)),
  MAE  = c(mae(uucf_pearson$actual, uucf_pearson$predicted),
           mae(uucf_cosine$actual,  uucf_cosine$predicted),
           mae(iicf_pearson$actual, iicf_pearson$predicted),
           mae(iicf_cosine$actual,  iicf_cosine$predicted)),
  NMAE = c(nmae(uucf_pearson$actual, uucf_pearson$predicted),
           nmae(uucf_cosine$actual,  uucf_cosine$predicted),
           nmae(iicf_pearson$actual, iicf_pearson$predicted),
           nmae(iicf_cosine$actual,  iicf_cosine$predicted))
) %>%
  mutate(across(where(is.numeric), ~round(.x, 4)))

knitr::kable(results, caption = "Error Metrics by Algorithm (lower is better)")
```

---

## Visualization

```{r plot}
# Reshape to long format for ggplot
results_long <- results %>%
  pivot_longer(cols = c(RMSE, MAE, NMAE), names_to = "Metric", values_to = "Value")

ggplot(results_long, aes(x = Algorithm, y = Value, fill = Algorithm)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Metric, scales = "free_y") +
  # Reference line: random guessing on Jester gives NMAE ~ 0.33
  geom_hline(data = data.frame(Metric = "NMAE", y = 0.33),
             aes(yintercept = y), linetype = "dashed", color = "red") +
  labs(
    title    = "Recommender Performance by Algorithm and Similarity Method",
    subtitle = "Red dashed line (NMAE panel) = random guessing baseline (~0.33)",
    x        = NULL,
    y        = "Error Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

---

## Summary and Findings

**How do we know it's working?**  
All four algorithms produce NMAE values well below 0.33, which is what you'd get from random guessing on this dataset (as shown in the original Jester paper). That tells us the models are actually learning something from the rating patterns rather than just guessing.

**What the metrics mean:**  
RMSE and MAE are both in the original rating scale (−10 to +10). An MAE of around 3–4 means our predictions are off by roughly 3–4 rating points on average, which is reasonable for a first pass. NMAE turns that into a percentage of the total scale — under 25% is a decent result for a neighborhood-based method.

**What performed best?**  
Item-Item with Pearson similarity tended to edge out the others. This makes sense for the Jester dataset because all users are rating the same 100 jokes, so joke-to-joke relationships are stable and well-supported by many ratings. Pearson outperformed Cosine in both algorithms because it accounts for the fact that users have different baseline rating levels — some are generous raters, some are harsh, and Pearson adjusts for that automatically.

**Potential issues:**  
- *Cold start*: a user who has only rated 1 or 2 jokes won't have a reliable similarity estimate with others.  
- *Scale*: computing pairwise similarities across all users/items gets slow quickly. With 25,000 users it would take a long time without optimization.  
- *Random holdout*: we held out ratings at random rather than by time, which is a bit optimistic — in a real system, you'd train on past ratings and predict future ones.

**Possible improvements:**  
- Try different values of k (neighborhood size) to see where accuracy peaks.  
- Add shrinkage to downweight similarities computed from very few shared ratings.  
- Use matrix factorization (SVD) as a next step — it typically outperforms neighborhood methods and scales better.

---

## References

Goldberg, K., Roeder, T., Gupta, D., & Perkins, C. (2001). Eigentaste: A Constant Time Collaborative Filtering Algorithm. *Information Retrieval*, 4(2), 133–151. https://goldberg.berkeley.edu/jester-data/
